---
title: "Shark Ear Hair"
author: "C Mull"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

## Shark Ear Hair
This script is for the analysis of ear hair density in 6 innner ear components (lag, sac, utr, mn_pos, mn_ant, and mn_avg) and how it corresponds to hearing sensitivty across a range of frequencies. First the data will be loaded and the response variables (Thr_int, PAL_int) and densities will be standardized (scaled and centered) to allow for comparison across structures.

```{r libraries, echo=FALSE}
#library(plyr)
library(tidyverse)
library(readxl)
library(brms)
library(brmstools)
```


```{r load data, echo=FALSE}

#Data is in an excel file on different sheets for each species (Rig, School, Carpet). Load each individually and bind.

#load species sheets
earHairRig <- read_excel("DataMaster_HearingVSHairs_adjusted.xlsx", 
    sheet = "PaperRig") %>% 
  mutate(species = "Rig")

earHairSchool <- read_excel("DataMaster_HearingVSHairs_adjusted.xlsx", 
    sheet = "PaperSchool") %>% 
  mutate(species = "School")

earHairCarpet <- read_excel("DataMaster_HearingVSHairs_adjusted.xlsx", 
    sheet = "PaperCarpet") %>% 
  mutate(species = "Carpet")

#bind into longform data and clean
earHair <- earHairRig %>% 
  bind_rows(earHairSchool) %>% 
  bind_rows(earHairCarpet)

#scale and center variables
earHair <- earHair %>% 
  filter(subject != "Rig5") %>% 
  mutate(Thr_int_scale = scale(Thr_int, center = T, scale = T),
         PAL_int_scale = scale(PAL_int, center = T, scale = T),
         #Accel_int_raw_scale = scale(Accel_int_raw, center = T, scale = T), #two decimal points in this column?
         lag_density_scale = scale(`lag_density/10,000`, center = T, scale = T),
         sac_density_scale = scale(`sac_density/10,000`, center = T, scale = T),
         utr_density_scale = scale(`utr_density/10,000`, center = T, scale = T),
         mn_pos_density_scale = scale(`mn_pos_density/10,000`, center = T, scale = T),
         mn_ant_density_scale = scale(`mn_ant_density/10,000`, center = T, scale = T),
         mn_avg_density_scale = scale(`mn_average_density/10,000`, center = T, scale = T)) %>% 
  dplyr::rename(lag_density = `lag_density/10,000`,
                sac_density = `sac_density/10,000`,
                utr_density = `utr_density/10,000`,
                mn_ant_density = `mn_ant_density/10,000`,
                mn_pos_density = `mn_pos_density/10,000`,
                mn_avg_density = `mn_average_density/10,000`)

```


# Exploratory Plots

Visualize how the chosen thresholds (Thr_int, PAL_int) vary across frequencies and species need to plot threshold over Fr for each species (line plots with error at each frequency). First need summary stats by groupings

```{r helper function for group sd, echo=FALSE}

data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
 return(data_sum)
}
```

```{r summaries of thresholds}
df_Thr <- data_summary(earHair, varname="Thr_int_scale", 
                    groupnames=c("species", "frequency"))
head(df_Thr)

df_PAL <- data_summary(earHair, varname="PAL_int_scale", 
                    groupnames=c("species", "frequency"))
head(df_PAL)
```

### Thresholds over fequencies by species

```{r plot of thresholds over frequencies by species, echo=FALSE}
ThrvFreq <- ggplot(df_Thr, aes(x=frequency, y=Thr_int_scale, color = species)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin=Thr_int_scale-sd, ymax=Thr_int_scale+sd), width=.2,
                 position=position_dodge(0.1))+
  theme(legend.position = "none")

PALvFreq <- ggplot(df_PAL, aes(x=frequency, y=PAL_int_scale, color = species)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin=PAL_int_scale-sd, ymax=PAL_int_scale+sd), width=.2,
                 position=position_dodge(0.05))

ggpubr::ggarrange(ThrvFreq, PALvFreq, ncol = 2, nrow = 1, common.legend = T, legend = "bottom")
```

This looks like it doesn't matter if we choose Thr or PAL as the response across frequencies as their scaled values are almost identical. Notably there is a drop off in sensitivity across all species at the highest frequencies, while carpet sharks exhibit a decline in sensitivity after 300 MHz.


### Densities across inner ear structures by species

```{r summaries of densities}

#select species

earHair %>% 
  dplyr::select(species, subject, "lag"= "lag_density", "sac"="sac_density", "utr"="utr_density", "mn_pos"= "mn_pos_density", "mn_ant"= "mn_ant_density", "mn_avg"= "mn_avg_density") %>% 
  distinct() %>% 
  pivot_longer(cols = lag:mn_avg, values_to = "hair_density", names_to = "structure") %>% 
  dplyr::group_by(species, structure) %>% 
  dplyr::summarise(density = mean(hair_density, na.rm=T),
                   sd = sd(hair_density, na.rm=T)) %>% 
  ggplot(aes(x=structure, y = density, fill = species))+
  geom_bar(stat="identity", color="black", 
           position=position_dodge()) +
  geom_errorbar(aes(ymin=density-sd, ymax=density+sd), width=.2,
                 position=position_dodge(.9))

```

The biggest differences looks to be in the density of hair cells in lag followed by sac and utr.




## Model

We fit a Bayesian linear mixed-effects model predicting hearing sensitivity (PAL_int_scale) from hair cell densities across 4 inner ear structures (lagena, saccule, utricle, and macula neglecta) and a smooth term for frequency. Species is treated as a fixed effect rather than a random grouping because only 3 species are present — too few levels to reliably estimate a random-effect variance. Individual subjects are retained as a random intercept nested within species to account for repeated measures across frequencies.

Hair cell density predictors across structures are likely correlated, so we also fit separate single-structure models and use LOO-IC to compare their predictive performance.

```{r fit models, cache=TRUE}

refit <- FALSE  # set to TRUE to refit models, FALSE to load saved versions

if (refit) {

  # Shared priors
  my_priors <- c(
    prior(normal(0, 1), class = "b"),
    prior(normal(0, 2), class = "Intercept"),
    prior(exponential(1), class = "sd"),
    prior(exponential(1), class = "sigma")
  )

  # Combined model: all three structures
  fit_combined <- brm(
    PAL_int_scale ~ lag_density_scale + sac_density_scale + utr_density_scale +
      species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )

  # Single-structure models for LOO comparison
  fit_lag <- brm(
    PAL_int_scale ~ lag_density_scale + species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )

  fit_sac <- brm(
    PAL_int_scale ~ sac_density_scale + species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )

  fit_utr <- brm(
    PAL_int_scale ~ utr_density_scale + species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )
  
   fit_mn_pos <- brm(
    PAL_int_scale ~ mn_pos_density_scale + species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )
   
   fit_mn_ant <- brm(
    PAL_int_scale ~ mn_ant_density_scale + species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )
   
  fit_mn_avg <- brm(
    PAL_int_scale ~ mn_avg_density_scale + species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  ) 

  saveRDS(fit_combined, "fit_combined.rds")
  saveRDS(fit_lag,      "fit_lag.rds")
  saveRDS(fit_sac,      "fit_sac.rds")
  saveRDS(fit_utr,      "fit_utr.rds")
  saveRDS(fit_mn_pos,   "fit_mn_pos.rds")
  saveRDS(fit_mn_ant,   "fit_mn_ant.rds")
  saveRDS(fit_mn_avg,   "fit_mn_avg.rds")

} else {

  fit_combined <- readRDS("fit_combined.rds")
  fit_lag      <- readRDS("fit_lag.rds")
  fit_sac      <- readRDS("fit_sac.rds")
  fit_utr      <- readRDS("fit_utr.rds")
  fit_mn_pos      <- readRDS("fit_mn_pos.rds")
  fit_mn_ant      <- readRDS("fit_mn_ant.rds")
  fit_mn_avg      <- readRDS("fit_mn_avg.rds")

}

```


```{r model diagnostics}

# Check for divergences — should be 0 after the fix
np_combined <- nuts_params(fit_combined)
cat("Divergent transitions (combined model):", sum(np_combined$Value[np_combined$Parameter == "divergent__"]), "\n")

# Full summary
summary(fit_combined)

```


```{r posterior predictive check}

# Overlay of observed vs. model-predicted distributions
pp_check(fit_combined, ndraws = 100)

```


```{r model plots}

# Trace plots and marginal posteriors for all parameters
plot(fit_combined)

# Non-linear frequency effect
plot(conditional_smooths(fit_combined))

# Conditional effects of each hair cell density predictor (marginalising over others)
plot(conditional_effects(fit_combined, effects = "lag_density_scale"))
plot(conditional_effects(fit_combined, effects = "sac_density_scale"))
plot(conditional_effects(fit_combined, effects = "utr_density_scale"))
plot(conditional_effects(fit_combined, effects = "mn_pos_density_scale"))
plot(conditional_effects(fit_combined, effects = "mn_ant_density_scale"))
plot(conditional_effects(fit_combined, effects = "mn_avg_density_scale"))

```


```{r loo comparison}

# Add LOO criterion to each model (uses existing MCMC draws — no refitting)
fit_combined <- add_criterion(fit_combined, "loo")
fit_lag      <- add_criterion(fit_lag,      "loo")
fit_sac      <- add_criterion(fit_sac,      "loo")
fit_utr      <- add_criterion(fit_utr,      "loo")
fit_mn_avg   <- add_criterion(fit_mn_avg,   "loo")
fit_mn_pos   <- add_criterion(fit_mn_pos,   "loo")
fit_mn_ant   <- add_criterion(fit_mn_ant,   "loo")

# Compare models — higher ELPD (lower LOOIC) is better
loo_compare(fit_combined, fit_lag, fit_sac, fit_utr, fit_mn_avg, fit_mn_pos, fit_mn_ant)

```

## Model selection and posterior interpretation

### LOO-IC model comparison

LOO-IC comparison ranked `fit_mn_avg` as the best-predicting model. However, the three macula neglecta models (`fit_mn_avg`, `fit_mn_pos`, `fit_mn_ant`) are statistically indistinguishable from one another (elpd_diff/SE ratios of 0.4 and 0.6 respectively, both well below the threshold of 2), reflecting the fact that posterior, anterior, and average macula neglecta density are measuring related aspects of the same structure. The lagena (`fit_lag`), saccule (`fit_sac`), and utricle (`fit_utr`) models perform somewhat worse but their differences also do not clearly exceed twice their standard errors. Only `fit_combined` is meaningfully worse than the top model (elpd_diff = −2.3, SE = 1.1, ratio ≈ 2.1), consistent with the earlier finding that combining correlated predictors reduces rather than improves predictive performance.

`fit_mn_avg` is taken as the primary model for interpretation, as average macula neglecta density summarises anterior and posterior regions jointly and offers the most parsimonious representation of that structure.

### Posterior distributions — fit_mn_avg

```{r posterior plot, fig.height=6}

library(tidybayes)
library(ggdist)

# Extract draws for parameters of interest
draws_mn_avg <- fit_mn_avg %>%
  spread_draws(b_mn_avg_density_scale, b_speciesRig, b_speciesSchool, sigma) %>%
  pivot_longer(cols = c(b_mn_avg_density_scale, b_speciesRig, b_speciesSchool, sigma),
               names_to = "parameter", values_to = "value") %>%
  mutate(parameter = recode(parameter,
    b_mn_avg_density_scale = "mn average density (scaled)",
    b_speciesRig           = "species: Rig vs Carpet",
    b_speciesSchool        = "species: School vs Carpet",
    sigma                  = "residual SD (sigma)"
  ))

ggplot(draws_mn_avg, aes(x = value, y = parameter, fill = parameter)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = mean_qi,
               normalize = "panels") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(x = "Posterior estimate", y = NULL,
       title = "Posterior distributions — fit_mn_avg",
       subtitle = "Point = mean, inner bar = 66% CI, outer bar = 95% CI. Dashed line at zero.") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

```

### Interpretation

**Macula neglecta average hair cell density** — the posterior mean is positive (β ≈ 0.13) and the 95% credible interval is close to but just spans zero (−0.01 to 0.26). This is the strongest density effect observed across all structures tested, and is suggestive of a positive relationship between macula neglecta hair cell density and hearing threshold (PAL_int). The direction — higher density associated with higher (worse) thresholds — warrants biological consideration and may in part reflect interspecific differences in both sensory morphology and hearing ability.

**Species differences** remain the most clearly supported signals in the model. Rig sharks have credibly higher hearing thresholds than Carpet sharks (the reference level; β ≈ 0.68, 95% CI: 0.34–1.01, P(β > 0) ≈ 99.9%). School sharks show a positive and credible offset (β ≈ 0.38, 95% CI: 0.07–0.70, P(β > 0) ≈ 99%).

**Residual variance** remains substantial (σ ≈ 0.67 on a standardised response), indicating that species identity and macula neglecta density together account for a moderate but incomplete proportion of variation in hearing sensitivity. The frequency spline captures additional structured variance not attributable to hair cell density.



## Species-specific audiograms

The model above assumes a single shared frequency response curve shifted additively per species. The exploratory plots suggest carpet sharks may have a distinct audiogram shape, particularly at higher frequencies. We therefore fit a model with species-specific frequency smooths (`by = species`) to test whether audiogram shape differs across species, and compare it to `fit_mn_avg` via LOO-IC.

```{r fit species-specific smooths, cache=TRUE}

if (refit) {

  fit_by_species <- brm(
    PAL_int_scale ~ mn_avg_density_scale + species +
      s(frequency, by = species, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )

  saveRDS(fit_by_species, "fit_by_species.rds")

} else {

  fit_by_species <- readRDS("fit_by_species.rds")

}

```


```{r species-smooth diagnostics}

np_by_species <- nuts_params(fit_by_species)
cat("Divergent transitions:", sum(np_by_species$Value[np_by_species$Parameter == "divergent__"]), "\n")

summary(fit_by_species)

```


```{r species-smooth loo comparison}

fit_by_species <- add_criterion(fit_by_species, "loo")
fit_mn_avg     <- add_criterion(fit_mn_avg,     "loo")

loo_compare(fit_by_species, fit_mn_avg)

```

A ratio of |elpd_diff / se_diff| > 2 indicates the species-specific smooth model is meaningfully better supported. If the ratio is below 2, the shared-curve model (`fit_mn_avg`) is preferred on parsimony grounds.

```{r species audiogram plot, fig.height=5}

# Plot species-specific frequency smooths with posterior uncertainty
conditions <- data.frame(species = c("Carpet", "Rig", "School"))
rownames(conditions) <- c("Carpet", "Rig", "School")

ce <- conditional_effects(
  fit_by_species,
  effects = "frequency:species",
  conditions = conditions,
  re_formula = NA
)

plot(ce, plot = FALSE)[[1]] +
  labs(x = "Frequency (Hz)", y = "Hearing threshold (PAL, scaled)",
       title = "Species-specific audiograms",
       subtitle = "Posterior mean ± 95% CI. Fitted from species-specific frequency smooths.",
       colour = "Species", fill = "Species") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

```

### Compare species senstitivities across frequencies

We will compare the species specific hearing thresholds across the range of frequencies using the estimated marginal means.


```{r pairwise contrasts by frequency, fig.height=5, fig.width=8}

library(emmeans)

# Marginal means per species at each recorded frequency
emm <- emmeans(fit_by_species, ~ species | frequency,
               at = list(frequency = sort(unique(earHair$frequency))),
               re_formula = NA)

# Tidy pairwise contrasts
pw <- as.data.frame(pairs(emm)) %>%
  dplyr::rename(lower = lower.HPD, upper = upper.HPD) %>%
  mutate(
    frequency = factor(frequency),
    credible  = !(lower < 0 & upper > 0),  # TRUE if 95% HPD excludes zero
    contrast  = recode(contrast,
      "Carpet - Rig"    = "Carpet vs Rig",
      "Carpet - School" = "Carpet vs School",
      "Rig - School"    = "Rig vs School"
    )
  )

ggplot(pw, aes(x = frequency, y = estimate, colour = credible)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_linerange(aes(ymin = lower, ymax = upper), linewidth = 0.8) +
  geom_point(size = 2.5) +
  facet_wrap(~ contrast, ncol = 3) +
  scale_colour_manual(
    values = c("FALSE" = "grey60", "TRUE" = "#185FA5"),
    labels = c("FALSE" = "CI includes zero", "TRUE" = "CI excludes zero"),
    name   = NULL
  ) +
  labs(
    x        = "Frequency (Hz)",
    y        = "Contrast estimate (scaled PAL threshold)",
    title    = "Pairwise species contrasts by frequency",
    subtitle = "Points = posterior median, bars = 95% HPD interval. Blue = credible difference."
  ) +
  theme_bw() +
  theme(
    panel.grid.minor  = element_blank(),
    legend.position   = "bottom",
    strip.background  = element_rect(fill = "grey95", colour = "grey80")
  )

```


Carpet sharks have less sensitive hearing across the highest frequencies. While Rig has more sensitive hearing than school shark 450 and 600 Hz, and likely 800 Hz with only a small portion of the HPD crossing zero.
