---
title: "Shark Ear Hair"
author: "C Mull"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

## Shark Ear Hair
This script is for the analysis of ear hair density in 6 innner ear components (lag, sac, utr, mn_pos, mn_ant, and mn_avg) and how it corresponds to hearing sensitivty across a range of frequencies. First the data will be loaded and the response variables (Thr_int, PAL_int) and densities will be standardized (scaled and centered) to allow for comparison across structures.

```{r libraries, echo=FALSE}
#library(plyr)
library(tidyverse)
library(readxl)
library(brms)
library(brmstools)
```


```{r load data, echo=FALSE}

#Data is in an excel file on different sheets for each species (Rig, School, Carpet). Load each individually and bind.

#load species sheets
earHairRig <- read_excel("DataMaster_HearingVSHairs_adjusted.xlsx", 
    sheet = "PaperRig") %>% 
  mutate(species = "Rig")

earHairSchool <- read_excel("DataMaster_HearingVSHairs_adjusted.xlsx", 
    sheet = "PaperSchool") %>% 
  mutate(species = "School")

earHairCarpet <- read_excel("DataMaster_HearingVSHairs_adjusted.xlsx", 
    sheet = "PaperCarpet") %>% 
  mutate(species = "Carpet")

#bind into longform data and clean
earHair <- earHairRig %>% 
  bind_rows(earHairSchool) %>% 
  bind_rows(earHairCarpet)

#scale and center variables
earHair <- earHair %>% 
  filter(subject != "Rig5") %>% 
  mutate(Thr_int_scale = scale(Thr_int, center = T, scale = T),
         PAL_int_scale = scale(PAL_int, center = T, scale = T),
         #Accel_int_raw_scale = scale(Accel_int_raw, center = T, scale = T), #two decimal points in this column?
         lag_density_scale = scale(`lag_density/10,000`, center = T, scale = T),
         sac_density_scale = scale(`sac_density/10,000`, center = T, scale = T),
         utr_density_scale = scale(`utr_density/10,000`, center = T, scale = T),
         mn_pos_density_scale = scale(`mn_pos_density/10,000`, center = T, scale = T),
         mn_ant_density_scale = scale(`mn_ant_density/10,000`, center = T, scale = T),
         mn_avg_density_scale = scale(`mn_average_density/10,000`, center = T, scale = T)) %>% 
  dplyr::rename(lag_density = `lag_density/10,000`,
                sac_density = `sac_density/10,000`,
                utr_density = `utr_density/10,000`,
                mn_ant_density = `mn_ant_density/10,000`,
                mn_pos_density = `mn_pos_density/10,000`,
                mn_avg_density = `mn_average_density/10,000`)

```


# Exploratory Plots

Visualize how the chosen thresholds (Thr_int, PAL_int) vary across frequencies and species need to plot threshold over Fr for each species (line plots with error at each frequency). First need summary stats by groupings

```{r helper function for group sd, echo=FALSE}

data_summary <- function(data, varname, groupnames){
  require(plyr)
  summary_func <- function(x, col){
    c(mean = mean(x[[col]], na.rm=TRUE),
      sd = sd(x[[col]], na.rm=TRUE))
  }
  data_sum<-ddply(data, groupnames, .fun=summary_func,
                  varname)
  data_sum <- rename(data_sum, c("mean" = varname))
 return(data_sum)
}
```

```{r summaries of thresholds}
df_Thr <- data_summary(earHair, varname="Thr_int_scale", 
                    groupnames=c("species", "frequency"))
head(df_Thr)

df_PAL <- data_summary(earHair, varname="PAL_int_scale", 
                    groupnames=c("species", "frequency"))
head(df_PAL)
```

### Thresholds over fequencies by species

```{r plot of thresholds over frequencies by species, echo=FALSE}
ThrvFreq <- ggplot(df_Thr, aes(x=frequency, y=Thr_int_scale, color = species)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin=Thr_int_scale-sd, ymax=Thr_int_scale+sd), width=.2,
                 position=position_dodge(0.1))+
  theme(legend.position = "none")

PALvFreq <- ggplot(df_PAL, aes(x=frequency, y=PAL_int_scale, color = species)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin=PAL_int_scale-sd, ymax=PAL_int_scale+sd), width=.2,
                 position=position_dodge(0.05))

ggpubr::ggarrange(ThrvFreq, PALvFreq, ncol = 2, nrow = 1, common.legend = T, legend = "bottom")
```

This looks like it doesn't matter if we choose Thr or PAL as the response across frequencies as their scaled values are almost identical. Notably there is a drop off in sensitivity across all species at the highest frequencies, while carpet sharks exhibit a decline in sensitivity after 300 MHz.


### Densities across inner ear structures by species

```{r summaries of densities}

#select species

earHair %>% 
  dplyr::select(species, subject, "lag"= "lag_density", "sac"="sac_density", "utr"="utr_density", "mn_pos"= "mn_pos_density", "mn_ant"= "mn_ant_density", "mn_avg"= "mn_avg_density") %>% 
  distinct() %>% 
  pivot_longer(cols = lag:mn_avg, values_to = "hair_density", names_to = "structure") %>% 
  dplyr::group_by(species, structure) %>% 
  dplyr::summarise(density = mean(hair_density, na.rm=T),
                   sd = sd(hair_density, na.rm=T)) %>% 
  ggplot(aes(x=structure, y = density, fill = species))+
  geom_bar(stat="identity", color="black", 
           position=position_dodge()) +
  geom_errorbar(aes(ymin=density-sd, ymax=density+sd), width=.2,
                 position=position_dodge(.9))

```

The biggest differences looks to be in the density of hair cells in lag followed by sac and utr.




## Model

We fit a Bayesian linear mixed-effects model predicting hearing sensitivity (PAL_int_scale) from hair cell densities across three inner ear structures (lagena, saccule, utricle) and a smooth term for frequency. Species is treated as a fixed effect rather than a random grouping because only 3 species are present — too few levels to reliably estimate a random-effect variance. Individual subjects are retained as a random intercept nested within species to account for repeated measures across frequencies.

Hair cell density predictors across structures are likely correlated, so we also fit separate single-structure models and use LOO-IC to compare their predictive performance.

```{r fit models, cache=TRUE}

refit <- FALSE  # set to TRUE to refit models, FALSE to load saved versions

if (refit) {

  # Shared priors
  my_priors <- c(
    prior(normal(0, 1), class = "b"),
    prior(normal(0, 2), class = "Intercept"),
    prior(exponential(1), class = "sd"),
    prior(exponential(1), class = "sigma")
  )

  # Combined model: all three structures
  fit_combined <- brm(
    PAL_int_scale ~ lag_density_scale + sac_density_scale + utr_density_scale +
      species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )

  # Single-structure models for LOO comparison
  fit_lag <- brm(
    PAL_int_scale ~ lag_density_scale + species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )

  fit_sac <- brm(
    PAL_int_scale ~ sac_density_scale + species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )

  fit_utr <- brm(
    PAL_int_scale ~ utr_density_scale + species + s(frequency, k = 8) + (1 | species:subject),
    data = earHair,
    family = gaussian(),
    prior = my_priors,
    chains = 4,
    iter = 10000,
    cores = 4,
    control = list(adapt_delta = 0.99, max_treedepth = 15)
  )

  saveRDS(fit_combined, "fit_combined.rds")
  saveRDS(fit_lag,      "fit_lag_v2.rds")
  saveRDS(fit_sac,      "fit_sac.rds")
  saveRDS(fit_utr,      "fit_utr.rds")

} else {

  fit_combined <- readRDS("fit_combined.rds")
  fit_lag      <- readRDS("fit_lag_v2.rds")
  fit_sac      <- readRDS("fit_sac.rds")
  fit_utr      <- readRDS("fit_utr.rds")

}

```


```{r model diagnostics}

# Check for divergences — should be 0 after the fix
np_combined <- nuts_params(fit_combined)
cat("Divergent transitions (combined model):", sum(np_combined$Value[np_combined$Parameter == "divergent__"]), "\n")

# Full summary
summary(fit_combined)

```


```{r posterior predictive check}

# Overlay of observed vs. model-predicted distributions
pp_check(fit_combined, ndraws = 100)

```


```{r model plots}

# Trace plots and marginal posteriors for all parameters
plot(fit_combined)

# Non-linear frequency effect
plot(conditional_smooths(fit_combined))

# Conditional effects of each hair cell density predictor (marginalising over others)
plot(conditional_effects(fit_combined, effects = "lag_density_scale"))
plot(conditional_effects(fit_combined, effects = "sac_density_scale"))
plot(conditional_effects(fit_combined, effects = "utr_density_scale"))

```


```{r loo comparison}

# Add LOO criterion to each model (uses existing MCMC draws — no refitting)
fit_combined <- add_criterion(fit_combined, "loo")
fit_lag      <- add_criterion(fit_lag,      "loo")
fit_sac      <- add_criterion(fit_sac,      "loo")
fit_utr      <- add_criterion(fit_utr,      "loo")

# Compare models — lower LOOIC is better
loo_compare(fit_combined, fit_lag, fit_sac, fit_utr)

```

## Model selection and posterior interpretation

### LOO-IC model comparison

LOO-IC comparison ranked `fit_utr` as the best-predicting model, with `fit_sac` statistically indistinguishable (elpd_diff = 0.0, SE = 1.0), `fit_lag` also not meaningfully different (elpd_diff = −0.3, SE = 1.0), and `fit_combined` marginally worse (elpd_diff = −1.5, SE = 0.7, ratio ≈ 2.1). Adding all three predictors simultaneously reduced predictive performance, likely because the hair cell density measures are correlated across structures, introducing noise rather than independent signal. `fit_utr` is taken as the primary model for interpretation.

### Posterior distributions — fit_utr

```{r posterior plot, fig.height=6}

library(tidybayes)
library(ggdist)

# Extract draws for parameters of interest
draws_utr <- fit_utr %>%
  spread_draws(b_utr_density_scale, b_speciesRig, b_speciesSchool, sigma) %>%
  pivot_longer(cols = c(b_utr_density_scale, b_speciesRig, b_speciesSchool, sigma),
               names_to = "parameter", values_to = "value") %>%
  mutate(parameter = recode(parameter,
    b_utr_density_scale = "utr hair cell density (scaled)",
    b_speciesRig        = "species: Rig vs Carpet",
    b_speciesSchool     = "species: School vs Carpet",
    sigma               = "residual SD (sigma)"
  ))

ggplot(draws_utr, aes(x = value, y = parameter, fill = parameter)) +
  stat_halfeye(.width = c(0.66, 0.95), point_interval = mean_qi,
               normalize = "panels") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(x = "Posterior estimate", y = NULL,
       title = "Posterior distributions — fit_utr",
       subtitle = "Point = mean, inner bar = 66% CI, outer bar = 95% CI. Dashed line at zero.") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

```

### Interpretation

**Utricle hair cell density** — the posterior mean is positive (β ≈ 0.08) but the 95% credible interval spans zero (−0.08 to 0.24), with approximately 85% of posterior mass above zero. This is suggestive of a weak positive relationship between utricle hair cell density and hearing threshold (PAL_int), but the evidence is insufficient to exclude a null effect. The direction of the effect — higher density associated with higher (worse) hearing thresholds — warrants biological scrutiny and may reflect interspecific confounding.

**Species differences** are the most clearly supported signals in the model. Rig sharks have credibly higher hearing thresholds than Carpet sharks (the reference level; β ≈ 0.52, 95% CI: 0.14–0.89, P(β > 0) ≈ 99.5%). School sharks show a positive but marginal offset (β ≈ 0.30, 95% CI: −0.06–0.67, P(β > 0) ≈ 95%).

**Residual variance** remains high (σ ≈ 0.67 on a standardised response), indicating that species identity and utricle hair cell density together explain only a moderate proportion of the variation in hearing sensitivity. Frequency (captured by the spline term) accounts for additional structured variance.

